Effectful Semantics for Substructural SSA

### 2023-11-13 pldi24/paper.tex
Printing monad

\begin{definition}[Submonad] 
  A \textbf{submonad} \(\mb{S}\) of a monad \(\mb{T}\) on \(\ms{Set}\) is a collection of sets such that, for all \(A\), \(\mb{S}\;A \subseteq \mb{T}\;A\), \(\forall a \in A, \eta\;a \in \mb{S}\;A\), and, for all \(s \in \mb{S}\;A\), \(f: A \to \mb{S}\;B\), \(\mb{bind}\;s\;f \in \mb{S}\;A\). We say that a submonad of \(\mb{T}\) is \textbf{Elgot} if \(\mb{T}\) is Elgot and, for all \(f: A \to \mb{S}(B + A)\), \(f^\dagger: A \to \mb{S}\;B\) where \(f^\dagger\) is computed as in \(\mb{T}\).
\end{definition}
Some important examples of submonads include:
\begin{itemize}
  \item The \textbf{unit submonad} \(\ms{pure}(\mb{T})\) of every monad \(\mb{T}\) with \(\ms{pure}(\mb{T})\;A = \{\eta_A\;a \mid a \in A\}\); similarly, \(\mb{T}\) is itself a submonad of \(\mb{T}\).
  \item Given submonads \(\mb{S}_i\) of \(\mb{T}\), their elementwise intersections \(\bigwedge_i\mb{S}_i = \lambda A. \bigcap_i \mb{S}_i\;A\) are submonads, forming a complete meet-semilattice with bottom element \(\ms{pure}(\mb{T})\) and top element \(\mb{T}\); this induces a complete lattice in the standard manner
  \item \todo{commuting absorbtive submonad of trace monad: example of...}
  \item \todo{central submonad. Go cite \cite{carette-central-2022}}
  \item \todo{deterministic, idempotent trace monad: relevant submonad}
  \item \todo{noneffectful trace monad: affine submonad}
  \item \todo{cap and centralize, cap and purify}
\end{itemize}
\TODO{a submonad induces a wide subcategory}

\TODO{in particular, these induce a substructural effectful category...}

\subsection{Heaps and Separation Logic}

@@

We'll start by considering the category \(\ms{HRel}\), which we will take to have morphisms of the form \(A \to \ms{Heap} \to \mc{P}^+(A \times \ms{Heap} \times \ms{Bytes}^* + \ms{Bytes}^{\leq \omega})\), which is the result of applying the state-transformer monad with state \(\ms{Heap}\) to the trace monad with the monoid of lists of bytestrings described in the previous section.

\TODO{say something better than "bytestrings"}

\TODO{what is a \(\ms{Heap}\)}

\TODO{Use name \ms{Traces}}

>>>>

\begin{definition}[Relevant, Affine Monad]
  A monad \(\mb{T}\) on \(\ms{Set}\) is \textbf{relevant} if its Kleisli
  category satisfies \(f;\ms{split} = \ms{split};f \ltimes f = \ms{split};f
  \rtimes f\), where \(\ms{split} = \upg{\lambda a. (a, a)}{}\). \(\mb{T}\) is
  \textbf{affine} if its Kleisli category satisfies \(f;\ms{drop} = \ms{drop}\),
  where \(\ms{drop} = \upg{\lambda a.()}{}\). A monad is \textbf{intuitionistic}
  if it is both relevant and affine.
\end{definition}

\begin{definition}[Submonad] 
  A \textbf{submonad} \(\mb{S}\) of a monad \(\mb{T}\) on \(\ms{Set}\) is a
  collection of sets such that, for all \(A\), \(\mb{S}\;A \subseteq
  \mb{T}\;A\), \(\forall a \in A, \eta\;a \in \mb{S}\;A\), and, for all \(s \in
  \mb{S}\;A\), \(f: A \to \mb{S}\;B\), \(\mb{bind}\;s\;f \in \mb{S}\;A\). We say
  that a submonad of \(\mb{T}\) is \textbf{Elgot} if \(\mb{T}\) is Elgot and,
  for all \(f: A \to \mb{S}(B + A)\), \(f^\dagger: A \to \mb{S}\;B\) where
  \(f^\dagger\) is computed as in \(\mb{T}\). Note that \(\mb{S}\) is also an
  (Elgot) monad in its own right, inheriting \(\mu, \eta\) and \(\cdot^\dagger\)
  from \(\mb{T}\).
\end{definition}
Some important examples of submonads include:
\begin{enumerate}
  \item The \textbf{unit submonad} \(\ms{pure}(\mb{T})\) of every monad
  \(\mb{T}\) with \(\ms{pure}(\mb{T})\;A = \{\eta_A\;a \mid a \in A\}\);
  similarly, \(\mb{T}\) is itself a submonad of \(\mb{T}\).
  \item Given submonads \(\mb{S}_i\) of \(\mb{T}\), their elementwise
  intersections \(\bigwedge_i\mb{S}_i = \lambda A. \bigcap_i \mb{S}_i\;A\) are
  submonads, forming a complete meet-semilattice with bottom element
  \(\ms{pure}(\mb{T})\) and top element \(\mb{T}\); this induces a complete
  lattice in the standard manner.
  \item In particular, the intersection of a relevant/affine/commutative
  submonad and another submonad is relevant/affine/commutative respectively.
  Since every commutative submonad induces a central subcategory of the Kleisli
  category of the parent monad, it follows that a triple of a relevant, affine,
  and commutative submonad make the Kleisli category of a monad \(\mb{T}\) into
  a substructural effectful category with all objects both affine and relevant.
  \item If \(C \subseteq Z(M)\) is submonoid of \(M\), then \(A \mapsto
  \mc{P}^+(A \times C + \mb{0}), \mc{P}^+(A \times C \times I) \subseteq
  \ms{Traces}\;\Theta\;A\) are submonads
  \item In particular, if \(M_C \subseteq Z(M)\) is a central submonoid of \(M\)
  such that \(\forall m \in M_C, \forall i \in I, m \cdot i = i\), then \(A
  \mapsto \mc{P}^+(A \times M_C + \mb{0}) \subseteq \ms{Traces}\;\Theta\;A\) is
  a submonad whose Kleisli category is a central subcategory of the Kleisli
  category of \(\ms{Traces}\;\Theta\;A\). Note \(M_C = \{1\}\) is always such a
  submonoid.
  \label{item:central-submonad}
  \item Similarly, if \(M_R \subseteq M_C\) is a submonoid such that \(\forall m
  \in M_R, m \cdot m = m\), then \(A \mapsto \{S \in \mc{P}^+(A \times M_R + I)
  \mid |S| = 1\} \subseteq \ms{Traces}\;\Theta\;A\) is a \textit{relevant}
  submonad. Note \(M_R = \{1\}\) is always such a submonoid.
  \label{item:relevant-submonad}
  \item Finally, \(A \mapsto \mc{P}^+(A \times \{1\} + \mb{0}) \subseteq
  \ms{Traces}\;\Theta\;A\) is an \textit{affine} submonad.
  \label{item:affine-submonad}
\end{enumerate}
In particular, \ref{item:central-submonad}, \ref{item:relevant-submonad},
\ref{item:affine-submonad} equip the Kleisli category of
\(\ms{Traces}\;\Theta\;A\) with the structure of a substructural effectful
category given choices of \(M_C\), \(M_R\).

\subsection{Heaps, Printing, and Separation Logic}

As a concrete example, we can now introduce the \textbf{printing monad}, which
is simply defined as \(\ms{Print}\;A \equiv \ms{Traces}\;\Sigma\), where
\(\Sigma: (\ms{Bytes}^*)^\omega \to \ms{Bytes}^{\leq \omega}\) denotes
concatenation of lists of bytestrings. This can be equipped with the structure
of a substructural effectful category by choosing \(M_C = M_R = \{[]\}\), giving
us an \isotopessa model supporting the instructions \(\ms{print} \in
\mc{I}^\varnothing_0(\ms{Bytes}, \mb{1})\) and \(\ms{rand} \in
\mc{I}^\taff_1(\mb{1}, A)\) for \(A\) nonempty, with standard semantics
\(\dnt{\ms{print}} = \lambda b.\{\iota_0 ((), [b])\}\), \(\dnt{\ms{rand}} =
\lambda (). \{\iota_0 (a, p[]) \mid a \in A\}\).

We can make things a little more interesting by adding a heap to the mix, defining our monad \(\ms{Comp} = \ms{StateT}\;\ms{Heap}\;\ms{Print}\), where \(\ms{Heap} = \nats \rightharpoonup \nats\) is simply a partial function. The Kleisli category of this monad can be equipped with the structure of a substructural effectful category by simply taking the affine/relevant morphisms to be the lifts of the affine/relevant morphisms in \(\ms{Set}_{\ms{Print}}\); as it inherits an Elgot structure from that on \(\ms{Set}_{\ms{Print}}\), we therefore have an \isotopessa model supporting the instructions \(\ms{get}: \mc{I}^\varnothing_0(\nats, \nats)\) and \(\ms{set}: \mc{I}^\varnothing_0(\nats \times \nats, ())\). We can assign semantics to \(\ms{set}\) in the standard fashion, with \(\dnt{\ms{set}} = \lambda (p, v)\;h. \{\iota_0((), [p \mapsto v]h, [])\}\). \(\ms{get}\) is a bit more tricky, since it is unclear what to do when we try to access uninitialized memory: one option is simply to return an arbitrary value, with \(\dnt{\ms{get}} = \lambda p\;h. \{\iota_0(v, []) \mid v = h\;p \lor p \notin h\}\); for now, this will do.

\TODO{segue to separation logic}

@@
